ProbabilityBoundsAnalysis.jl:\\ an arithmetic of probabilities and sets of probabilities

### 2021-09-30 paper/paper.tex
final draft

% JuliaCon proceedings template
\documentclass{juliacon}
\setcounter{page}{1}
\usepackage{caption}
\usepackage{csquotes}
\usepackage{amsmath}
%\bibliographystyle{juliacon}

@@

Probability bounds analysis combines interval arithmetic with probability theory, and provides a representation of sets of distributions in structures called probability boxes (p-boxes), which generalise both distribution functions and intervals. P-boxes generally return interval bounds on all probabilistic quantities, for example samples, cdfs, and probability measures are all intervals. This framework also allows for the comprehensive propagation of probabilities through calculations in a rigorous way, in a similar fashion that interval arithmetic does for sets. As such, ProbabilityBoundsAnalysis.jl gives a rigorous arithmetic of random variables, where both marginal (univariate) distributions and dependency information can be known, partially known or missing completely. We describe the main theoretical elements of probability bounds analysis, and provide a simplified implementation of the method in code snippets which can be readily evaluated in the Julia command terminal.

@@

\noindent Indeed it was Kolmogorov who originally asked about the result of an operation between two random variables without knowing their joint distribution. This was answered for the sum by Makarov \cite{makarov1982estimates}, who showed the result was a set of distributions and was able to provide bounds on this function. Sklar, Schweizer and Frank generalised this result to other (positive) binary operations \cite{frank1987best,schweizer2011probabilistic}. In this pursuit they created copulas, a general way to encode probabilistic dependence independently from marginals, and a now essential object used in probabilistic modelling. In his seminal dissertation \cite{williamson1989probabilistic}, Williamson described an algorithm for efficiently performing these arithmetic operations, which can give guaranteed bounds on probability distributions in terms of an upper and lower cdf. He called his method \textit{Probabilistic Arithmetic} and his sets of distributions \textit{Dependency Bounds}. Since then the method has been generalised \cite{ferson2015constructing,ferson1996whereof,ferson2004arithmetic} to most of the base binary and unary operations that would be present in a programming language. Probability boxes (p-boxes) are the name now given to these structures, and Probability Bounds Analysis (PBA) the name of the method.

The goal of PBA can be stated as \textbf{to compute guaranteed bounds on functions of random variables given only partial knowledge of the input probability distributions and their dependencies}. That is to compute with partial knowledge about the input joint distribution. Ideally all of the available information about random variables should be used, but no more than what actually is available.

The idea of bounding probability has a very long tradition throughout the history of probability theory. George Boole \cite{boole1854investigation, hailperin1986boole} used the notion of interval bounds on probability. Chebyshev \cite{chebyshev1874valeurs} described bounds on a distribution when only the mean and variance of the variable are known, and Markov \cite{markoff1900question} found bounds on a positive variable when only the mean is known.  Fréchet \cite{frechet1935generalisation} discovered how to bound joint distributions solely from knowing the marginal distributions, without making independence assumptions. Bounding probabilities has continued to the present day, culminating into the modern theory of Imprecise Probabilities \cite{walley1991statistical, klir2013uncertainty, troffaes2014lower, augustin2014introduction}.

Imprecise probability is effectively a generalisation of probability theory where uncertainty can be expressed about the probability measure. This is particularly relevant when information is scarce, unreliable, vague, conflicting or imprecise. In such cases defining a unique probability distribution is difficult. P-boxes are one of many ways to describe a set of distributions, others include: Dempster-Shafer structures \cite{dempster2008upper,shafer1976mathematical}, random sets \cite{molchanov2005theory}, possibility distributions \cite{zadeh1978fuzzy,dubois1988possibility, hose2019possibilistic} and lower previsions \cite{troffaes2014lower}. These structures were discovered independently, but are often synonymous and can be translated from one to another, with different degrees of generality. Imprecise probability links all these theories into one. For a comprehensive overview of the theory, and a for a formal description of uncertainty and information in terms of these structures, \cite{klir2013uncertainty} is recommended. In that sense PBA is a part of imprecise probabilities but provides a framework for computing with p-boxes.

@@

\noindent That is, a random variable is a member of a p-box if its cdf $F$ falls within the cdf bounds of the p-box $F(x) \in [\underline{F}(x), \overline{F}(x)]$ for all $x$, its moments are inside the interval moments of the p-box, and it belongs to a family of distribution functions (e.g. normal, uniform) considered by the p-box. Some of the constraints may be missing. For example, if the distribution family is unknown then the set is defined solely from the cdf and moment bounds. Such p-boxes are sometimes called non-parametric, since its members do not belong to any particular class of distribution. Some constraints may also be inferred from others. For example, the interval moments may be bounded from the cdf bounds, and cdf may be bounded from moment information (explored further in section \ref{section:Moments}). Figure \ref{fig:figure1} shows an example of a distribution function, interval and a p-box. The grey shaded region in the interval and p-box bounds a (potentially infinite) collection of probability distributions.

@@

Moreover, theoretical bounds on the mean and variance of an interval can be found [CITE SCOTT]:

@@

An important feature of probability bounds analysis is how distribution functions and p-boxes are represented. Generally, analytical solutions for cdfs are not readily available, for example the normal distribution's cdf can only be found by integrating the probability density, usually with quadrature. However even if the functions were available analytically, when the variables are used in an arithmetic operation the output distribution will not necessarily belong to the same family. Therefore we require a representation of these continuous functions which is not dependent on the distribution family (for example as used in polynomial chaos expansion), and ideally is robust in the sense that it can produce an interval error for the numerical representation. Such a representation of a probability is sometimes called an outer approximation, in the same fashion an interval is an outer approximation of a floating point number, and which contracts to the exact value as more computational resources is used.

We follow the representation first introduced by Williamson and Downs \cite{williamson1990probabilistic}, where a discrete upper and lower approximations of distribution functions are constructed using the inverse cdfs. Note that when considering inverse cdfs $\underline{F}^{-1}(p) \geq F^{-1}(p) \geq \overline{F}^{-1}(p)$, for $p \in [0,1]$. An outer approximation using two finite vectors\footnote{We begin vector indexing with $1$ as is done in the Julia language.} $u$ and $d$ of length $N$ is constructed by evaluating the inverses for uniformly spaced probability levels $p_{i} = \frac{i-1}{N}$ for $i = 1, ... , N+1$, e.g. $p_{i} = [0, 0.25, 0.5, 0.75, 1]$ for $N = 4$. The vectors $u$ and $d$ are defined as follows:

@@

A minimal p-box structure is as follows:

@@

@@

Other distributional p-boxes are constructed in a similar way. A evaluation of the cdf in terms of this representation is as follows:

@@

How the moment transformations are performed are not described here, however are available in \textit{MomentArithmetic.jl}\footnote{https://github.com/AnderGray/MomentArithmetic.jl at time of writing} \cite{ferson2021distribution}. Figure \ref{fig:figure4} shows an example of an operation between a precise distribution and an interval.

@@

In the case where a binary operation is performed between two distributions but whose dependence or correlation is unknown, then a single output distribution also cannot be defined. However a bounding p-box can be computed. This is the original question of Kolmogorov as described in the introduction. We call such operations between p-boxes where the marginals are fixed but no joint information is known \textit{Fréchet} operations, after the mathematician Maurice R. Fréchet, who first discovered how to bound joint distributions with fixed marginals \cite{frechet1935generalisation}. How operations between p-boxes with partial or no dependence information is known is described in section \ref{sec:pboxBinary}. Figure \ref{fig:figure5} illustrates a Fréchet operation between a uniform and a normal distribution.

@@

Outer approximations of precise distributions described in section \ref{sec:outer_approx} also lead naturally to a p-box. All probabilistic quantities of discretised distribution functions will be intervals, which shrink to precise quantities as the discretisation increases. In that sense, both distributions functions and p-boxes are p-boxes in ProbabilityBoundsAnalysis.jl.

@@

An example is robust bayesian analysis, where the analyst does not have enough information to reliably establish a single prior and likelihood function, but instead has a collection of priors and likelihoods, leading to a set of posteriors distributions, to which a p-box can be identified.

A method which is growing in popularity are inferential models \cite{martin2015inferential}. These generalise classical inference methods to account for limited statistical knowledge, and gives a statistical inference theory which is non-additive (as opposed to classical probability which is additive). These inference procedures, often based on random sets, give certain validity and performance guarantees. Inferential models can viewed from the perspective of bounding sets of distributions, and is closely linked to possibility theory \cite{liu2020inferential}. A p-box can be prescribed to such structures, leading to structures like confidence boxes or c-boxes \cite{fersoncomputing}, which follow the same arithmetic rules as p-boxes.

@@

A minimal structure, cdf, random sample and probability measure of a bivariate p-box is \footnote{https://github.com/AnderGray/BivariateCopulas.jl can be used for 2-copulas}:

@@

\subsection{Unary operations}
Piggybacking off \textit{IntervalArithmetic.jl}\footnote{https://github.com/JuliaIntervals/IntervalArithmetic.jl at time of writing} \cite{david_p_sanders_2021_5519761}, a p-box can be transformed through any unary operation $f: \mathbb{R} \rightarrow \mathbb{R}$ that has  an interval extension, i.e. accepts intervals. For this, we first transform the p-box into a Dempster-Shafer structure \cite{dempster2008upper,shafer1976mathematical}, propagate the structure through the operation with interval arithmetic, and transform it back to a p-box. A Dempster-Shafer structure (also called a belief function) is a finite collection of intervals, called focal elements $X$, with probabilities attached to each interval, called basic mass assignments $m$, which must sum to $1$. An example of a belief function with three focal elements is: $X = \{[0, 2], [0.5, 1.3], [1, 3]\}$ and $m = \{0.25, 0.5, 0.25\}$. Like a p-box, a Dempster-Shafer also bounds a set of probability distributions, using two set based functions called the \textit{belief} and \textit{plausibility}:

@@

The general function where the focal elements have non-equal mass is simple, but slightly lengthy and will not be described here. A function to perform a unary operation on a p-box is thus:

@@

@@

Williamson and Downs provide an algorithm to compute rigorous bounds on the above convolution, but only under independence (when $C = \Pi$). Around the same time, Yager \cite{yager1986arithmetic} described a similar style convolution on two belief functions under the assumption independence, the algorithm of which is very similar to that of Williamson and Downs. We therefore provide a generalised version of Yager's algorithm where any copula $C$ can be specified. Similar to a unary operation, a binary operation between two p-boxes can be performed by first transforming the bivariate p-box into a two dimensional belief function (which has interval boxes as focal elements), evaluating the operation on the focal elements with interval arithmetic, and transforming the resulting belief function (which will be univariate after the operation) back to a p-box. Converting a bivariate p-box into a 2D belief function is similar to that of the 1D case, however this time the mass assignment $m$ will be determined by the copula $C$. For two univariate focal elements $X$ and $Y$ with probability widths $p_{x} = [p_{x1}, p_{x2}]$ and $p_{y} = [p_{y1}, p_{y2}]$ (i.e. their vertical width in the left Figure \ref{fig:figure2}, with their 1D mass assignment being $p_{x2} - p_{x1}$), the 2D mass assignment on the interval box $X \times Y$ will be:

@@

Again the general function for converting a belief function with non-equal masses is required. The 3rd argument \textit{num\_steps} is the desired discretisation of the output p-box, in a condensation step. The interval operation in the $\sigma$ convolution performs the operation on all combinations of focal elements, so for example if each p-box has 200 steps, the resulting p-box will have $200 \times 200 = 40,000$ steps, which must be reduced in an outer approximation. Again, the general algorithm is simple but lengthy and will be omitted.

@@

@@

\section*{Acknowledgements}

>>>>

% JuliaCon proceedings template
\documentclass{juliacon}
\setcounter{page}{1}
\usepackage{caption}
\usepackage{csquotes}
\usepackage{amsmath}
\usepackage{xcolor}
%\bibliographystyle{juliacon}

@@

Probability bounds analysis combines interval arithmetic with probability theory, and provides a representation of sets of distributions in structures called probability boxes (p-boxes). These which generalise both distribution functions and intervals. P-boxes generally return interval bounds on all probabilistic quantities, for example samples, cdfs, and probability measures are all intervals. This framework also allows for the comprehensive propagation of probabilities through calculations in a rigorous way, in a similar fashion that interval arithmetic does for sets. As such, \texttt{ProbabilityBoundsAnalysis.jl} gives a rigorous arithmetic of random variables, where both marginal (univariate) distributions and dependency information can be known, partially known or missing completely. We describe the main theoretical elements of probability bounds analysis, and provide a simplified implementation of the method in code snippets which can be readily evaluated in the Julia command terminal.

@@

\noindent Indeed it was Kolmogorov who originally asked about the result of an operation between two random variables without knowing their joint distribution. This was answered for the sum by Makarov \cite{makarov1982estimates}, who showed the result was a set of distributions and was able to provide bounds on this function. Sklar, Schweizer and Frank generalised this result to other (positive) binary operations \cite{frank1987best,schweizer2011probabilistic}. In this pursuit they created copulas, a general way to encode probabilistic dependence independently from marginals. These are now an essential object used in probabilistic modelling. In his dissertation \cite{williamson1989probabilistic}, Williamson described an algorithm for efficiently performing these arithmetic operations, which can give guaranteed bounds on probability distributions in terms of an upper and lower cdf. He called his method \textit{Probabilistic Arithmetic} and his sets of distributions \textit{Dependency Bounds}. Since then the method has been generalised \cite{ferson2015constructing,ferson1996whereof,ferson2004arithmetic} to most of the base binary and unary operations that would be present in a programming language. Probability boxes (p-boxes) are the name now given to these structures

The goal of \texttt{ProbabilityBoundsAnalysis.jl}\footnote{https://github.com/AnderGray/ProbabilityBoundsAnalysis.jl} is to \textbf{compute guaranteed bounds on functions of random variables given only partial knowledge of the input probability distributions and their dependencies}. That is to compute with partial knowledge about the input joint distribution. Ideally, all of the available information about random variables should be used, but no more than what actually is available.

The idea of bounding probability has a long tradition throughout the history of probability theory. George Boole \cite{boole1854investigation, hailperin1986boole} used the notion of interval bounds on probability. Chebyshev \cite{chebyshev1874valeurs} described bounds on a distribution when only the mean and variance of the variable are known, and Markov \cite{markoff1900question} found bounds on a positive variable when only the mean is known.  Fréchet \cite{frechet1935generalisation} discovered how to bound joint distributions solely from knowing the marginal distributions, without making independence assumptions. Bounding probabilities has continued to the present day, culminating into the modern theory of Imprecise Probabilities \cite{walley1991statistical, klir2013uncertainty, troffaes2014lower, augustin2014introduction}.

Imprecise probability is effectively a generalisation of probability theory where uncertainty can be expressed about the probability measure. This is particularly relevant when information is scarce, unreliable, vague, conflicting or imprecise. In such cases defining a unique probability distribution is difficult. P-boxes are one of many ways to describe a set of distributions, others include: Dempster-Shafer structures \cite{dempster2008upper,shafer1976mathematical}, random sets \cite{molchanov2005theory}, possibility distributions \cite{zadeh1978fuzzy,dubois1988possibility, hose2019possibilistic} and lower previsions \cite{troffaes2014lower}. These structures were discovered independently, but are often synonymous and can be translated from one to another, with different degrees of generality. Imprecise probability links all these theories into one. For a comprehensive overview of the theory, and a for a formal description of uncertainty and information in terms of these structures, \cite{klir2013uncertainty} is recommended.

@@

\noindent That is, a random variable is a member of a p-box if its cdf $F$ falls within the cdf bounds of the p-box $F(x) \in [\underline{F}(x), \overline{F}(x)]$ for all $x$, its moments are inside the interval moments of the p-box, and it belongs to a family of distribution functions (e.g. normal, uniform) considered by the p-box. Some of the constraints may be missing. For example, if the distribution family is unknown then the set is defined solely from the cdf and moment bounds. Such p-boxes are sometimes called non-parametric, since its members do not belong to any particular class of distribution. Some constraints may also be inferred from others. For example, the interval moments may be bounded from the cdf bounds, and cdf may be bounded from moment information (explored further in section \ref{section:Moments}). Figure\footnote{Scripts for all figures can be found in /examples/JuliaCon} \ref{fig:figure1} shows an example of a distribution function, interval and a p-box. The grey shaded region in the interval and p-box bounds a (potentially infinite) collection of probability distributions.

@@

Moreover, theoretical bounds on the mean and variance of an interval can be found:

@@

An important feature of probability bounds analysis is how distribution functions and p-boxes are represented. Generally, analytical solutions for cdfs are not readily available, for example the normal distribution's cdf can only be found by integrating the probability density, usually with quadrature. However, even if the functions were available analytically, when the variables are used in an arithmetic operation the output distribution will not necessarily belong to the same family. Therefore, we require a representation of these continuous functions which is not dependent on the distribution family (for example as used in polynomial chaos expansion), and ideally is robust in the sense that it can produce an interval error for the numerical representation. Such a representation of a probability is sometimes called an outer approximation, in the same fashion an interval is an outer approximation of a floating point number, which contracts to the exact value as more computational resources are used.

We follow the representation first introduced by Williamson and Downs \cite{williamson1990probabilistic}, where a discrete upper and lower approximations of distribution functions are constructed using the inverse cdfs. Note that when considering inverse cdfs $\underline{F}^{-1}(p) \geq F^{-1}(p) \geq \overline{F}^{-1}(p)$, for $p \in [0,1]$. An outer approximation using two finite vectors\footnote{We begin vector indexing with $1$ as is done in the Julia language.} $u$ and $d$ of length $N$ is constructed by evaluating the inverse cdfs for uniformly spaced probability levels $p_{i} = \frac{i-1}{N}$ for $i = 1, ... , N+1$, e.g. $p_{i} = [0, 0.25, 0.5, 0.75, 1]$ for $N = 4$. The vectors $u$ and $d$ are defined as follows:

@@

A minimal\footnote{The code for the minimal structure can be found in examples/minimal} p-box structure is as follows:

@@

    b1 = quantile.(Normal(m.lo, std.lo), ps)
    b2 = quantile.(Normal(m.lo, std.hi), ps)
    b3 = quantile.(Normal(m.lo, std.hi), ps)
    b4 = quantile.(Normal(m.hi, std.hi), ps)

    u = min.(b1, b2, b3, b4)[1:end-1]
    d = max.(b1, b2, b3, b4)[2:end]

\iffalse
\begin{lstlisting}[language = Julia]
  using Distributions
  const steps = 200

@@

    return pbox(u, d, m, std^2, "normal")
  end
\end{lstlisting}
\fi
Other distributional p-boxes are constructed in a similar way. A evaluation of the cdf in terms of this representation is as follows:

@@

How the moment transformations are performed are not described here, however they are available in \texttt{MomentArithmetic.jl}\footnote{https://github.com/AnderGray/MomentArithmetic.jl} \cite{ferson2021distribution}. Figure \ref{fig:figure4} shows an example of an operation between a precise distribution and an interval.

@@

In the case where a binary operation is performed between two distributions but whose dependence or correlation is unknown, a single output distribution also cannot be defined. However a bounding p-box can be computed. This is the original question of Kolmogorov as described in the introduction. We call such operations between p-boxes where the marginals are fixed but no joint information is known \textit{Fréchet} operations, after the mathematician Maurice R. Fréchet, who first discovered how to bound joint distributions with fixed marginals \cite{frechet1935generalisation}. How operations between p-boxes with partial or no dependence information is known is described in section \ref{sec:pboxBinary}. Figure \ref{fig:figure5} illustrates a Fréchet operation between a uniform and a normal distribution.

@@

Outer approximations of precise distributions described in section \ref{sec:outer_approx} also lead naturally to a p-box. All probabilistic quantities of discretised distribution functions will be intervals, which shrink to precise quantities as the discretisation increases. In that sense, both distributions functions and p-boxes are p-boxes in \texttt{ProbabilityBoundsAnalysis.jl}.

@@

An example is robust Bayesian analysis, where the analyst does not have enough information to reliably establish a single prior and likelihood function, but instead has a collection of priors and likelihoods, leading to a set of posteriors distributions, to which a p-box can be identified.

A method which is growing in popularity are inferential models \cite{martin2015inferential}. These generalise classical inference methods to account for limited statistical knowledge, and give a statistical inference theory which is non-additive (as opposed to classical probability which is additive). These inference procedures, often based on random sets, give certain validity and performance guarantees. Inferential models can viewed from the perspective of bounding sets of distributions, and is closely linked to possibility theory \cite{liu2020inferential}. A p-box can be prescribed to such structures, leading to structures like confidence boxes or c-boxes \cite{fersoncomputing}, which follow the same arithmetic rules as p-boxes.

@@

A minimal structure, cdf, random sample and probability measure of a bivariate p-box is \footnote{https://github.com/AnderGray/BivariateCopulas.jl}:

@@

\subsection{Unary operations}
Piggybacking off \texttt{IntervalArithmetic.jl}\footnote{https://github.com/JuliaIntervals/IntervalArithmetic.jl} \cite{david_p_sanders_2021_5519761}, a p-box can be transformed through any unary operation $f: \mathbb{R} \rightarrow \mathbb{R}$ that has  an interval extension, i.e. accepts intervals. For this, we first transform the p-box into a Dempster-Shafer structure \cite{dempster2008upper,shafer1976mathematical}, propagate the structure through the operation with interval arithmetic, and transform it back to a p-box. A Dempster-Shafer structure (also called a belief function) is a finite collection of intervals, called focal elements $X$, with probabilities attached to each interval, called basic mass assignments $m$, which must sum to $1$. An example of a belief function with three focal elements is: $X = \{[0, 2], [0.5, 1.3], [1, 3]\}$ and $m = \{0.25, 0.5, 0.25\}$. Like a p-box, a Dempster-Shafer also bounds a set of probability distributions, using two set based functions called the \textit{belief} and \textit{plausibility}:

@@

The general function where the focal elements have non-equal mass is simple, but slightly lengthy, and will not be described here. A function to perform a unary operation on a p-box is thus:

@@

  \centering
  \includegraphics[width=.25\textwidth]{../examples/JuliaCon/fig9/fig9_pbox1.pdf}\hfill
  \includegraphics[width=.25\textwidth]{../examples/JuliaCon/fig9/fig9_pbox2.pdf}\hfill
  \includegraphics[width=.25\textwidth]{../examples/JuliaCon/fig9/fig9_pbox3.pdf}\hfill
  \includegraphics[width=.25\textwidth]{../examples/JuliaCon/fig9/fig9_pbox4.pdf}

  \includegraphics[width=.25\textwidth]{../examples/JuliaCon/fig9/fig9_pbox5.pdf}\hfill
  \includegraphics[width=.25\textwidth]{../examples/JuliaCon/fig9/fig9_pbox6.pdf}\hfill
  \includegraphics[width=.25\textwidth]{../examples/JuliaCon/fig9/fig9_pbox7.pdf}\hfill
  \includegraphics[width=.25\textwidth]{../examples/JuliaCon/fig9/fig9_pbox8.pdf}

  \caption{Top left shows two initial p-boxes $X, Y \sim U([1,2], 4)$, and others show results of binary operations on these two p-boxes with different dependencies. Each colour represents a different copula. The colour code is $C = \{$ {\color{purple} $W$} (opposite), {\color{green} $\Phi_{-0.5}$ }(correlation -0.5), {\color{red} $\Pi$} (independence), {\color{orange} $\Phi_{0.5}$} (correlation 0.5), {\color{blue} $M$}(perfect) $\}$}
  \label{fig:figure9}

@@

Williamson and Downs provide an algorithm to compute rigorous bounds on the above convolution, but only under independence (when $C = \Pi$). Around the same time, Yager \cite{yager1986arithmetic} described a similar style convolution on two belief functions under the assumption independence, which is very similar to that of Williamson and Downs'. We therefore provide a generalised version of Yager's algorithm where any copula $C$ can be specified. Similar to a unary operation, a binary operation between two p-boxes can be performed by first transforming the bivariate p-box into a two dimensional belief function (which has interval boxes as focal elements), evaluating the operation on the focal elements with interval arithmetic, and transforming the resulting belief function (which will be univariate after the operation) back to a p-box. Converting a bivariate p-box into a 2D belief function is similar to that of the 1D case, however this time the mass assignment $m$ will be determined by the copula $C$. For two univariate focal elements $X$ and $Y$ with probability widths $p_{x} = [p_{x1}, p_{x2}]$ and $p_{y} = [p_{y1}, p_{y2}]$ (i.e. their vertical width in Figure \ref{fig:figure2}, with their 1D mass assignment being $p_{x2} - p_{x1}$), the 2D mass assignment of the interval box $X \times Y$ will be:

@@

The above $\sigma$ convolution will work for any two input p-boxes, with any copula $C$, and binary operation $op$ which has an interval extension. Again the general function for converting a belief function with non-equal masses is required. The 3rd argument \textit{num\_steps} is the desired discretisation of the output p-box, in a condensation step. The interval operation in the $\sigma$ convolution performs the operation on all combinations of focal elements, so for example if each p-box has 200 steps, the resulting p-box will have $200 \times 200 = 40,000$ steps, which must be reduced in an outer approximation. Again, the general algorithm is simple but lengthy and will be omitted. Figure \ref{fig:figure9} shows examples of $\sigma$ convolutions for different operators and copulas.

@@

\textit{ProbabilityBoundsAnalysis.jl} also allows for binary operations to be performed between p-boxes when only the lower bound of the copula is known $\underline{C}$. That is, uncertainty can be placed on the dependence of two distributions, as well as the marginals. The resulting p-box of the binary operation will bound all distributions with copulas $C$ which are pointwise greater than the specified copula $\underline{C} \leq C$. These operations are based on two more convolutions from Probabilistic Metric Spaces \cite{schweizer2011probabilistic}. For a non-decreasing binary operator $L: \mathbb{R}^+\times \mathbb{R}^+ \to \mathbb{R}^+$, two random variables $X$ and $Y$ with distribution functions $F_X$ and $F_Y$, and with copula lower bound $\underline{C}$:

\begin{equation*}\label{taupbox}
  \tau_{\underline{C}_{XY},L}(z) = \sup_{L(x,y)=z} [\underline{C}(F_{X}(x), F_Y(y))],
\end{equation*}
\begin{equation*}\label{rhopbox}
  \rho_{\underline{C}_{XY},L}(z) = \inf_{L(x,y)=z} [\underline{C}^d(F_{X}(x), F_Y(y))],
\end{equation*}

where $C^d$ is the copula dual of $C$, $C^d(u,v) = u + v - C(u,v)$. The $\tau$ and $\rho$ convolutions each compute a cdf bound which together characterise the resulting p-box: $[\underline{F}_{Z}(z), \overline{F}_{Z}(z)] = [\tau(z), \rho(z)]$. The p-box resulting from a $\tau$ and $\rho$ convolution will bound all $\sigma$ convolutions which have copula $C$ which is greater than $\underline{C}$. For example, setting $\underline{C} = W$ (the lower bound on all copulas) will propagate all copulas, which is complete uncertainty about the dependence, i.e. a \textit{Fréchet} operation. Setting $\underline{C} = \Pi$ will propagate copulas more positive than independence, otherwise known as \textit{positive quadrant dependence} \cite{nelsen2007introduction}. Setting $\underline{C} = M$ (the upper bound on all copulas) will only propagate $M$, and will be equal to a $\sigma$ convolution.

The $\tau$ and $\rho$ convolutions can be efficiently computed in terms of the inverse representation using in \textit{ProbabilityBoundsAnalysis}:

\begin{equation*}\label{taupbox2}
  \tau^{-1}_{\underline{C}_{XY},L}(p) = \sup_{\underline{C}(u,v) = p} [L(F^{-1}_{X}(u), F^{-1}_Y(v))],
\end{equation*}
\begin{equation*}\label{rhopbox2}
  \rho^{-1}_{\underline{C}_{XY},L}(p) = \inf_{\underline{C}^{d}(u,v) = p} [L(F^{-1}_{X}(u), F^{-1}_Y(v))],
\end{equation*}

where the operation $L$ is performed on the $u$ and $d$ vectors of the p-boxes, and $\sup_{\underline{C}(u,v) = p}$ and $\inf_{\underline{C}^{d}(u,v) = p}$ is solved for each probability level.

Note that $L$ must be non-decreasing function (for example $+$ or $\times$), but may be extended to non-increasing functions (for example $-$ or $\div$) by first transforming one of the variables, and performing a non-decreasing operation. For example, $X - Y$ can be evaluated with sum: $X + (-Y)$. Note however, the copula $C_{XY}$ must also be transformed to $C_{X,-Y}$. This transformation however is quite simple \cite{nelsen2007introduction}:

\begin{equation*}\label{rotation}
  C_{X, -Y}(u,v) = u - C_{XY}(u, 1 - v).
\end{equation*}

Note that when using the above copula transformation, the bound on $\underline{C}_{XY}$ will be flipped to $\overline{C}_{X, -Y}$, and $\tau_{M, -}$ and $\rho_{M,-}$ give the \textit{Fréchet} convolution.

@@

\section*{Acknowledgements}
We would like to thank our friends and colleagues at the Risk Institute who have supported \texttt{ProbabilityBoundsAnalysis.jl} from its birth, and through their contributions and discussions have made this software possible: Marco De Angelis, Nick Gray, Alex Wimbush and Dominic Calleja. We also thank Dominic and Marco's detailed review of this paper.

We are also specially grateful to the researchers at JuliaReach and JuliaIntervals, who through their interest, contributions, discussions and suggestions have greatly improved the software: Marcello Forets, Christian Schilling, Luis Benet and David P. Sanders.
